---
title: "STAT 331 Final Project"
author: "Krishna Prem Pasumarthy & Islam Amin"
date: "`r format(Sys.Date(), format = '%B %d, %Y')`"
output:
  pdf_document:
    number_sections: yes
  html_document:
    df_print: paged
---

\section{Summary}




\section{Descriptive Statistics}

First, take a look at summary statistics of the \texttt{fhsd} dataset.

```{r echo=FALSE}
fhsd <- read.csv('fhs.csv')

# summary statistics on explanatory variables
summary(fhsd)
```

Then take a look at \texttt{chdrisk} grouped by \texttt{sex} as well as \texttt{chdrisk} grouped by \texttt{cursmoke}.

```{r,echo=FALSE}
library(kableExtra)
by(fhsd$chdrisk, fhsd$sex, summary)
by(fhsd$chdrisk, fhsd$cursmoke, summary)
```


Now take a look at pair plots of all numeric explanatory variates i.e. variates excluding response variate \texttt{chdrisk} and logical variates such as \texttt{cursmoke}. 

```{r echo=FALSE}



# pairs(~ totchol + ldlc, data = fhsd)
# pairs(~ diabp + totchol, data = fhsd)
# pair plots for continuous variables
pairs(~ totchol + age + sysbp + diabp + cigpday + bmi + 
        heartrte + glucose + hdlc + ldlc, 
      col = "blue",                                         # Change color
      pch = 18,                                            # Change shape of points
      main = "Pair Plots of Continuous Variables",
      data = fhsd)
```

From the pair plots, we can observe a strong correlation between low density lipoprotein cholesterol and serum total cholestrol. This correlation could be explained by the fact that there could be a relationship between
the amount 

Now take a look at the VIFs of these variates.

```{r echo=FALSE}
# design matrix excluding intercept
X <- model.matrix(lm(chdrisk ~ . -  1, data = fhsd))
# remove linearly dependent column (sexMale = 1 - sexFemale)
X <- X[,-1]
# calculate vif
vif <- diag(solve(cor(X)))

vif
```

\section{Candidate Models}


\section{Model Diagnostics}


\section{Model Selection}


\section{Discussion}



